Solving the maximum independant set problem with einsum networks

### 2021-06-02 paper/paper.tex
update code

\section{Computing independence polynomial}
The independence polynomial~\cite{Ferrin2014, Harvey2017} of a graph $G$ is defined as
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$, and $\alpha(G)$ is the maximum independent set size.
\blue{here, I preassumed a user knows tensor network, add more details later.}
If we map a graph into an einsum network, as shown in \Fig{fig:einsummapping}, by placing a rank one tensor of size $2$ at each vertex
\begin{equation}
    W(x)_{s_i} = \left(\begin{matrix}
        1 \\
        x
    \end{matrix}\right)_{s_i},
\end{equation}
and a rank two tensor of size $2 \times 2$ at each edge
\begin{equation}
    B_{s_i s_j} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right)_{s_is_j},
\end{equation}
the contraction of this einsum network generates the independence polynomial %, we denote this contraction as \texttt{a,b,c,d,ab,ad,bc,bd,cd,ce->}.
\begin{equation}
    I(G, x) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j}.
\end{equation}
Here, the einsum enumerates all possible configurations and accumulates the product of tensor elements to the output.
The product over vertex tensors provides a factor $x^k$, where $k=\sum_i s_i$ is the set size.
While the product over edge tensors provides a factor $0$ if the configuration is not an independent set, otherwise $1$.
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \centerline{\includegraphics[width=0.8\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig1.pdf}}
        \caption{}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
    \centerline{\includegraphics[width=0.8\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig2.pdf}}
        \caption{}
    \end{subfigure}
    \caption{Mapping a graph to an einsum network.}\label{fig:einsummapping}
\end{figure*}
By mapping the problem to einsum, one can use some existing algorithms to decrease the complexity of the problem.
Especially those techiniques developed in recent quantum circuit simulations for finding the optimal exactly contraction order of a tensor network ~\cite{Gray2021,Pan2021}.
With these algorithms, one can easily find a good contraction order that reduces the complexity to a quantity related to the minimum treewidth of the graph.~\cite{Markov2008}
Such mapping also makes it possible to utilize faster BLAS functions.
Before contracting the tensor network and evaluating this polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A ring without additive inverse is called a semiring, and commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relation must hold for arbituary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} * a = \mymathbb{0}
\end{align*}

\subsection{Symbolic computing}
To evaluate this polynomial, one can use the symbolic computing directly.
One can store a polynomial of $x$ as a vector of factors.
We define the algebra between the polynomial expansions $a$ of order $k_a$ and $b$ of order $k_b$ as
\begin{align}
    \begin{split}
    a \oplus b &= \{a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}\},\\
    a \odot b &= \{a_0 + b_0, a_1b_0 + a_0b_1, \ldots, a_{k_a} b_{k_b}\},\\
    \mymathbb{0} &= \{\},\\
    \mymathbb{1} &= \{1\},
    \end{split}
\end{align}
and insert them to the original tensor network contraction algorithm.
In the program, the multiplication can be evaluated efficiently with the convolution theorem.
However, this method still suffers from a space overhead that propotional to the maximum independant set size.

@@

\section{Computing maximum independent set size and its corresponding degeneracy and configurations}
Obtaining the maximum independent set size and its degeneracy is much easier. Let $x=\infty$, then the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_k \infty^{\alpha(G)},
\end{equation}
where the lower orders terms disappear automatically. We can define a new algebra as
\begin{align}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}\\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}
\end{split}
\end{align}
In the program, we only store the power $x$ and the corresponding factor $a_x$ that initialized to $1$. This is exactly the tropical tensor network~\cite{Liu2021} used in solving spin glass ground states.
Here if we are only interested in obtaining $\alpha(G)$, we can only consider the powers, which corresponds to the tropical algebra~\cite{Maclagan2015}.

>>>>

\section{Three approaches to computing independence polynomial}
The independence polynomial~\cite{Ferrin2014, Harvey2017} of a graph $G$ is defined as
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$, and $\alpha(G)$ is the maximum independent set size.
\blue{here, I preassumed a user knows tensor network, add more details later.}
If we map a graph into an einsum network, as shown in \Fig{fig:einsummapping}, by placing a rank one tensor of size $2$ at each vertex
\begin{equation}
    W(x)_{s_i} = \left(\begin{matrix}
        1 \\
        x
    \end{matrix}\right)_{s_i},
\end{equation}
and a rank two tensor of size $2 \times 2$ at each edge
\begin{equation}
    B_{s_i s_j} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right)_{s_is_j},
\end{equation}
the contraction of this einsum network generates the independence polynomial %, we denote this contraction as \texttt{a,b,c,d,ab,ad,bc,bd,cd,ce->}.
\begin{equation}
    I(G, x) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j}.
\end{equation}
Here, the einsum enumerates all possible configurations and accumulates the product of tensor elements to the output.
The product over vertex tensors provides a factor $x^k$, where $k=\sum_i s_i$ is the set size.
While the product over edge tensors provides a factor $0$ if the configuration is not an independent set, otherwise $1$.
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \centerline{\includegraphics[width=0.8\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig1.pdf}}
        \caption{}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
    \centerline{\includegraphics[width=0.8\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig2.pdf}}
        \caption{}
    \end{subfigure}
    \caption{Mapping a graph to an einsum network.}\label{fig:einsummapping}
\end{figure*}
By mapping the problem to einsum, one can use some existing algorithms to decrease the complexity of the problem.
Especially those techiniques developed in recent quantum circuit simulations for finding the optimal exactly contraction order of a tensor network ~\cite{Gray2021,Pan2021}.
With these algorithms, one can easily find a good contraction order that reduces the complexity to a quantity related to the minimum treewidth of the graph.~\cite{Markov2008}
Such mapping also makes it possible to utilize faster BLAS functions.
Before contracting the tensor network and evaluating this polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A ring without additive inverse is called a semiring, and commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relation must hold for arbituary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} * a = \mymathbb{0}
\end{align*}

\subsection{Symbolic computing}
To evaluate this polynomial, one can use the symbolic computing directly, that is, storing a polynomial of $x$ as a vector of factors.
We define the algebra between the polynomials $a$ of order $k_a$ and $b$ of order $k_b$ as
\begin{align}
    \begin{split}
    a \oplus b &= \{a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}\},\\
    a \odot b &= \{a_0 + b_0, a_1b_0 + a_0b_1, \ldots, a_{k_a} b_{k_b}\},\\
    \mymathbb{0} &= \{\},\\
    \mymathbb{1} &= \{1\}.
    \end{split}
\end{align}
The variable $x$ now is represented as a vector $\{0, 1\}$.
Then we insert them to the original tensor network contraction algorithm, the final result is the exact representation of the independence polynomial.
In the program, the multiplication can be evaluated efficiently with the convolution theorem.
However, this method suffers from a space overhead that propotional to the maximum independant set size because each polynomial requires a vector of such size to store.

@@

\section{Computing maximum independent set size and its corresponding degeneracy and configurations}
Obtaining the maximum independent set size and its degeneracy is computational more efficient. Let $x=\infty$, then the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_k \infty^{\alpha(G)},
\end{equation}
where the lower orders terms disappear automatically. We can define a new algebra as
\begin{align}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}\\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}
\end{split}
\end{align}
In the program, we only store the power $x$ and the corresponding factor $a_x$ that initialized to $1$.
This algebra is consistent with the one we derived in ~\cite{Liu2021} that uses the tropical tensor network solving spin glass ground states.
Here if we are only interested in obtaining $\alpha(G)$ and omit the factors, the algebra of $x$ corresponds to the tropical algebra~\cite{Maclagan2015,Moore2011}.
